---
title: "Coding Sample: Regression Analysis and Model Diagnostics in R"
format:
  pdf:
    toc: false
    number-sections: false
    geometry: margin=1in
fontsize: 11pt
---

**Repository:** <https://github.com/syp4137/regression-analysis-coding-sample.git>

## Project Overview

This coding sample demonstrates a structured regression workflow in R, including data inspection, exploratory analysis, linear model fitting, nested model comparison, and regression diagnostics. I selected this sample because it reflects how I approach applied statistical analysis: organizing data, choosing interpretable models, evaluating fit, and clearly communicating results.

-   Language: R
-   Dataset: Duncan occupational prestige dataset
-   Focus: exploratory analysis, linear regression, model comparison, diagnostics, interpretation

## Data and Objective

Examine how occupational prestige is associated with education and income, and assess whether occupational type improves model fit.

## Exploratory Analysis

I began by inspecting the variables and summarizing the data structure. Initial exploration suggested that prestige was positively associated with education and income, motivating a linear modeling approach.

```{r}
library(car)  # for Duncan data and diagnostic plots

data(Duncan)

str(Duncan)
summary(Duncan)

scatterplotMatrix(~ income + education + prestige, data = Duncan, 
                  groups = Duncan$type, 
                  smooth = FALSE)

boxplot(prestige ~ type, data = Duncan,
        xlab = "Occupation Type", ylab = "Prestige")
```

## Model Fitting and Comparison

I first fit a baseline linear model using education and income, then added occupational type as a categorical predictor. To compare the two nested models, I used a partial F-test along with adjusted $R^2$ and residual standard error.

```{r}
m1 <- lm(prestige ~ education + income, data = Duncan)
m2 <- lm(prestige ~ education + income + type, data = Duncan)

summary(m1)
summary(m2)
anova(m1, m2)
```

The expanded model improved fit substantially, with a significant partial F-test and better adjusted $R^2$. This suggests that occupational type adds explanatory value beyond education and income alone.

## Diagnostic Checks

After selecting the preferred model, I examined studentized residuals to assess model assumptions. The residual distribution was broadly reasonable, though a few observations in the upper tail suggested mild departures from normality.

```{r fig.height=5, fig.width=5}
rs <- rstudent(m2)
summary(rs)

qqPlot(m2, main = "QQ Plot of Studentized Residuals")
densityPlot(rs, main = "Studentized Residuals")
```

Overall, the diagnostics did not suggest a severe violation of assumptions, though a small number of observations warranted attention.

## Key Takeaways

This sample highlights the kind of work I most enjoy: turning a structured statistical question into a clear analytical workflow, from data inspection through model comparison and diagnostics. It reflects the careful, interpretable, and reproducible approach I aim to bring to research-oriented data work.
